Accurate evaluation of dislocation tractions

### 2021-01-14 src/wip/DanielCelisGarza/Accurate_evaluation_of_dislocation_tractions/paper/main.tex
Edit 1 in the bag.

\begin{abstract}
    %% Text of abstract
    Dislocations generate tractions on the surface of a finite volume. Traditionally the dislocation tractions are integrated numerically over a finite element face to obtain the nodal forces; as required when using superposition to couple discrete dislocation plasticity and the finite element method. Here we implement analytic finite element nodal forces obtained using closed form expressions for the traction integrals. We compare the behaviour of the errors which arise when using numerical integration, provide insight into how and why they occur and give recommendations on avoiding numerical issues when implementing the analytic solution. Evaluating the tractions analytically in a large scale discrete dislocation plasticity simulations of a microcantilever showed a clear difference in the simulated load-displacement curve while the dislocation structure was unchanged.
\end{abstract}

@@

Coupling discrete dislocation dynamics (DDD) to the finite element method (FEM) allows micromechanical tests to be simulated \cite{Groh2009}. Virtual experiments are essential for interpreting experimental data and relating the measured mechanical response to dislocation mechanisms \cite{0965-0393-6-6-007,doi:10.1080/14786430500341250,tarleton2015discrete,YU2018}. One of the most popular methods for coupling DDD and FEM is the superposition method \cite{superposition_scheme0,superposition_scheme1,superposition_scheme2}.
%
    \includegraphics[width=\linewidth]{fem_ddd.pdf}
    \caption[Superposition Model for DDD-FEM coupling.]{The superposition used to couple DDD and FEM. The volume $V$ is bounded by a surface $S = S_{T} \cup S_{U}$ and contains a dislocation ensemble and is subjected to tractions $\vec{T}$ on $S_{T}$ and $\vec{u}$ on $S_{u}$. First, the traction, $\vec{\tilde{T}}$, and displacement, $\vec{\tilde{U}}$, fields due to the dislocations in the infinite domain (DDD) are evaluated on the boundaries $S_{T}$ and $S_{U}$ respectively. Then an elastic boundary value problem can be solved with FEM to calculate the corrective elastic fields required to satisfy the boundary conditions $\vec{\hat{T}} = \vec{T} - \vec{\tilde{T}}$ and $\vec{\hat{u}} = \vec{U} - \vec{\tilde{U}}$.}
    \label{f:superposition_scheme}
%        
The superposition scheme works by decomposing the problem into separate DDD and FE problems as in \cref{f:superposition_scheme}. A linear-elastic solid $V$ bounded by a surface $S$ is subjected to traction boundary conditions, $\vec{T}$, on $S_{T}$ and displacement boundary conditions, $\vec{U}$, on $S_{U}$. The ($\tilde{~}$) fields are those generates by the dislocations in an infinite solid and are obtained by evaluating analytic solutions in a DDD simulation \cite{Cai2006}. Formally the dislocation field satisfies
%

@@

     & \left.
    \begin{array}{l}
        \nabla\cdot{\vec{\tilde{\sigma}}} = 0                    \\
        \vec{\tilde{\sigma}} = \vec{C}:\tilde{\vec{\varepsilon}} \\
        \tilde{\vec{\varepsilon}}=\frac{1}{2}\left(\nabla\tilde{\vec{u}}+(\nabla\tilde{\vec{u}})^T\right)
    \end{array}
    \right\}\textrm{in V}                                                   \\
    %
     & \tilde{\vec{\sigma}}\cdot\vec{n} = \tilde{\vec{T}} \textrm{ on } S_T \\
    %
     & \left.
    \begin{array}{l}
        \tilde{\vec{u}} = \tilde{\vec{U}} \quad t>0 \\
        \tilde{\vec{u}} = \vec{0} \quad t=0
        \label{eq:utildebc}
    \end{array}
    \right\} \textrm{ on } S_U.

@@

%
As the dislocation fields do not vanish on $S$ the dislocations load the volume by generating tractions, $\tilde{\vec{T}}$, on $S_T$ and displacements, $\tilde{\vec{U}}$, on $S_U$ this additional loading deforms $V$ generating an additional ``image'' stress which the dislocations then feel. Therefore corrective ($\hat{~}$) fields must be superimposed to satisfy the desired boundary conditions. The corrective field which accounts for both the applied and image stress is obtained numerically by solving the elastic boundary value problem
%

@@

     & \left.
    \begin{array}{l}
        \nabla\cdot{\vec{\hat{\sigma}}} = 0                  \\
        \vec{\hat{\sigma}} = \vec{C}:\hat{\vec{\varepsilon}} \\
        \hat{\vec{\varepsilon}}=\frac{1}{2}\left(\nabla\hat{\vec{u}}+(\nabla\hat{\vec{u}})^{T}\right)
    \end{array}
    \right\}\textrm{in V}                                                                            \\
    %
     & \hat{\vec{\sigma}}\cdot\vec{n} = \vec{T} - \tilde{\vec{T}} \textrm{ on } S_T\label{eq:thatbc} \\
    %
     & \hat{\vec{u}} = \vec{U} - \tilde{\vec{U}} \textrm{ on } S_U \label{eq:uhatbc}

@@

%
once the solution to both problems are known, then their superposition solves the desired mixed boundary value problem:
%

@@

     & \left.
    \begin{array}{l}
        \nabla\cdot\vec{\sigma}=\nabla\cdot\left(\vec{\hat{\sigma}} +\vec{\tilde{\sigma}}\right) = 0                                                              \\
        \vec{\sigma} =\vec{\hat{\sigma}}+\vec{\tilde{\sigma}}= \vec{C}:\left(\hat{\vec{\varepsilon}}+\tilde{\vec{\varepsilon}}\right)=\vec{C}:{\vec{\varepsilon}} \\
        \vec{\varepsilon}=\hat{\vec{\varepsilon}}+\tilde{\vec{\varepsilon}}=\frac{1}{2}\left(\nabla(\hat{\vec{u}}+\tilde{\vec{u}})+\left(\nabla(\hat{\vec{u}}+\tilde{\vec{u}})\right)^T\right) =\frac{1}{2}\left(\nabla\vec{u}+(\nabla\vec{u})^{T}\right)
    \end{array}
    \right\}\textrm{in V}                                    \\
    %
     & \vec{\sigma}\cdot\vec{n} = \vec{T}  \textrm{ on } S_T \\
    %
     & \left.
    \begin{array}{l}
        \vec{u} = \vec{U} \quad t>0 \\
        \vec{u} = \vec{0} \quad t=0
    \end{array}
    \right\} \textrm{ on } S_U
    \label{eq:ubc}

@@

The method is not without problems however. As a dislocation segment moves closer to the surface, its ($\tilde{~}$) field diverges and starts causing numerical problems \cite{boundary_problems_in_dd}. This can be partially solved by using a non-singular formulation such as that proposed by \citet{Cai2006} however the gradients in the dislocation field are difficult to accurately capture when a dislocation approaches $S$. Another problem with this method is the computational cost when simulating a heterogeneous solid as it requires calculating polarisation stresses due to the difference in the elastic constants in a secondary phase \cite{superposition_scheme0,boundary_problems_in_dd,ddd_precipitate}. A modified superposition scheme \cite{ODay2004} can overcome this by dividing the problem into separate DDD/FEM problems coupled through 1 elastic FE problem. This requires accurate evaluation of $\tilde{T}$ on the domain boundaries which can only be captured with a fine FE mesh. This in turn will increase the computational cost. In order to simulate polycrystalline or composite materials using superposition requires methods to accurately evaluate both $\tilde{\vec{U}}$ and $\tilde{\vec{T}}$. The displacements can be evaluated analytically \cite{ddd_disp} and this paper investigates the evaluation of the dislocation tractions analytically.

The relative simplicity of the superposition method has made it a popular choice for coupling DDD and FEM since all it requires is the evaluation of FE nodal forces and displacements on the boundary. Furthermore, analytic expressions for the stress field produced by a finite straight dislocation line segment, has allowed \citet{Queyreau} to use a non-singular formulation to obtain a closed-form solution for the tractions generated by a segment on the surface of a finite element. They used integration by parts giving rise to a number of recursion relations that can be used to construct the full solution from a small set of seed functions.

The force exerted by a dislocation ensemble on a node $a$ belonging to element $e$ is given by,
%
\begin{align}
    \vec{F}_{a} = \int_{S_{e}} \left[\tns{\tilde{\sigma}}(\vec{x}) \cdot \vec{n}\right] N_{a}(\vec{x})\; \mathrm{d}S_{e}\,,
    \label{eq:ddd_fem_force}
\end{align}
%
where $\mathrm{d}S_{e}$ is the surface of element $e$ with surface area $S_{e}$ and $N_{a}$ is the finite element shape function for node $a$ which is bi-linear in this paper. In the parent element (shown in \cref{f:2d_gaussian_quad}) the shape functions are,

\begin{align}
    \label{eq:shape_function}
    N_{1} & = \dfrac{1}{4}(1-s_1)(1-s_2)             \\
    N_{2} & = \dfrac{1}{4}(1+s_1)(1-s_2)\nonumber    \\
    N_{3} & = \dfrac{1}{4}(1+s_1)(1+s_2)\nonumber    \\
    N_{4} & = \dfrac{1}{4}(1-s_1)(1+s_2)\nonumber\,.
\end{align}

Usually Gauss quadrature is used to evaluate the surface integral in \cref{eq:ddd_fem_force} numerically. The 1D Gauss quadrature is,
%
\begin{align}
    \label{eq:gauss_leg}
    \int\limits_{-1}^{1} f(s)\;\mathrm{d}s \approx \sum\limits_{i=1}^{n} w_{i} f(s_{i}) \quad \textrm{where}\quad
    %
    w_{i} = \dfrac{2}{\left(1-s_{i}^{2}\right) \left[P_{n}^{'}\left(s_{i}\right)\right]^{2}},
\end{align}
%
is the weighting of the Gauss point $s_{i}$ which is the $i\textsuperscript{th}$ root of $P_{n}$; the $n\textsuperscript{th}$ Legendre polynomial normalised so $P_{n}(1) = 1$. $P_{n}^{'}$ is the first order derivative of $P_{n}$.

The method is very accurate for functions that can be approximated by polynomials, but unsuitable for functions with singularities and near singularities \cite{gauss_leg, gauss_leg_sing}. We use a non-singular formulation for the forces \cite{Cai2006}, but the singularity is avoided by adding a small cut off radius to account for the dislocation core. However if the integration point falls close to or within the dislocation core, numerical integration can still generate large errors.

For rectangular surface elements we must transform from the parent element in $(s_1,\,s_2)$ in \cref{eq:gauss_leg} to the real element coordinate system $(x,y,z)$ shown in \cref{f:force_lin_rect}. Evaluating \cref{eq:ddd_fem_force} in the parent element and mapping to the real element gives the force on node $a$,
    \label{eq:gauss_leg_ddd_fem_force}
    \vec{F}_{a} \approx \sum\limits_{i=1}^{Q} w_{i} \sum\limits_{j=1}^{Q} w_{j} [\tns{\tilde{\sigma}}(r_{i}, s_{j})\cdot \vec{n}] N_{a}(r_{i}, s_{j}) \det(\mtx{J})\,.

@@

Where the sum is over the $Q$ quadrature points and $J_{ij}=dx_{i}/ds_{j}$ is the element Jacobian defining the transformation from $(s_1,\,s_2) \mapsto (x,\,y)$; as the real element is rectangular with surface area $S_e$, $\det(\mtx{J})=S_e/4$. \Cref{f:2d_gaussian_quad} contains examples of how the Gauss points are distributed on the surface.
    \includegraphics[width=\linewidth]{2d_gaussian_quad.pdf}
    \caption{Examples of 2D Gauss-Legendre quadrature of the parent element with $Q = 1,\, 2,\, 3,\, 4$. The point size represents the weight $w$ of the integration point. The parent elements are centred at the origin and $s_1,\, s_2 \in [-1,\,1]$. We use an anticlockwise node numbering scheme.}
    \label{f:2d_gaussian_quad}
    \includegraphics[width=0.8\linewidth]{force_calc_linear_rectangle.pdf}
    \caption{Diagram of the parametric line integrals solved by \citet{Queyreau} to find the forces on linear rectangular surface elements.}
    \label{f:force_lin_rect}

>>>>

\begin{abstract}
    %% Text of abstract
    Dislocations generate tractions on the surface of a finite volume. Traditionally the dislocation tractions are integrated numerically over a finite element face to obtain the nodal forces; as required when using superposition to couple discrete dislocation plasticity and the finite element method. Here we implement analytic finite element nodal forces obtained using closed form expressions for the traction integrals. We compare the behaviour of the errors which arise when using numerical integration, provide insight into how and why they occur and give recommendations on avoiding numerical issues when implementing the analytic solution. Comparisons to infinite-domain solutions and a very simple unloaded simulation shows the drastic differences between the two methods.
\end{abstract}

@@

Coupling discrete dislocation dynamics (DDD) to the finite element method (FEM) allows micromechanical tests to be simulated with explicit dislocation interactions \cite{Groh2009}. Virtual experiments are essential for interpreting experimental data and relating the measured mechanical response to dislocation mechanisms \cite{0965-0393-6-6-007,doi:10.1080/14786430500341250,tarleton2015discrete,YU2018}. One of the most popular and simplest methods for coupling DDD and FEM is the superposition method \cite{superposition_scheme0,superposition_scheme1,superposition_scheme2}, which works by decomposing the problem into separate DDD and FE problems as illustrated in \cref{f:superposition_scheme}.
%
\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{fem_ddd.pdf}
    \caption[Superposition Model for DDD-FEM coupling.]{The superposition used to couple DDD and FEM. The volume $V$ is bounded by a surface $S = S_{T} \cup S_{U}$ and contains a dislocation ensemble and is subjected to tractions $\vec{T}$ on $S_{T}$ and $\vec{u}$ on $S_{u}$. First, the traction, $\vec{\tilde{T}}$, and displacement, $\vec{\tilde{U}}$, fields due to the dislocations in the infinite domain (DDD) are evaluated on the boundaries $S_{T}$ and $S_{U}$ respectively. Then an elastic boundary value problem can be solved with FEM to calculate the corrective elastic fields required to satisfy the boundary conditions $\vec{\hat{T}} = \vec{T} - \vec{\tilde{T}}$ and $\vec{\hat{u}} = \vec{U} - \vec{\tilde{U}}$.}
    \label{f:superposition_scheme}
\end{figure}

Specifically, a linear-elastic solid $V$ bounded by a surface $S$ is subjected to traction boundary conditions, $\vec{T}$, on $S_{T}$ and displacement boundary conditions, $\vec{U}$, on $S_{U}$. The ($\tilde{~}$) fields are those generated by the dislocations in an infinite solid and in our case are obtained by evaluating analytic solutions in a DDD simulation \cite{Cai2006}. Formally, the dislocation field satisfies,
\begin{align}
     & \left.
    \begin{array}{l}
        \nabla \cdot \tns{\tilde{\sigma}} = \vec{0}                \\
        \tns{\tilde{\sigma}} = \tns{C} : \tns{\tilde{\varepsilon}} \\
        \tns{\tilde{\varepsilon}} = \frac{1}{2} \left( \nabla \vec{\tilde{u}} + (\nabla \vec{\tilde{u}})^T \right)
    \end{array}
    \right\} \quad \textrm{in V}                                                   \\
    %
     & \tilde{\tns{\sigma}} \cdot \vec{n} = \tilde{\vec{T}} \quad \textrm{on } S_T \\
    %
     & \left.
    \begin{array}{l}
        \vec{\tilde{u}} = \vec{\tilde{u}}, \quad t>0 \\
        \vec{\tilde{u}} = \vec{0}, \quad t=0
        \label{eq:utildebc}
    \end{array}
    \right\} \quad \textrm{on } S_U \,.
\end{align}
%
As the dislocation fields do not vanish on $S$, the dislocations load the volume by generating tractions, $\tilde{\vec{T}}$, on $S_T$ and displacements, $\vec{\tilde{u}}$, on $S_U$. This additional loading deforms $V$, generating an additional ``image'' stress which the dislocations then feel. Therefore corrective ($\hat{~}$) fields must be superimposed to satisfy the desired boundary conditions. The corrective field which accounts for both the applied and image stress is obtained numerically by solving the elastic boundary value problem,
%
\begin{align}
     & \left.
    \begin{array}{l}
        \nabla \cdot \tns{\hat{\sigma}} = \vec{0}              \\
        \tns{\hat{\sigma}} = \tns{C} : \tns{\hat{\varepsilon}} \\
        \tns{\hat{\varepsilon}} = \frac{1}{2} \left( \nabla \vec{\hat{u}} + (\nabla \vec{\hat{u}})^{T} \right)
    \end{array}\right\} \quad \textrm{in V}                                                  \\
    %
     & \hat{\tns{\sigma}}\cdot\vec{n} = \vec{T} - \tilde{\vec{T}} \quad \textrm{on } S_T \label{eq:thatbc} \\
    %
     & \vec{\hat{u}} = \vec{U} - \vec{\tilde{u}} \quad \textrm{on } S_U \,. \label{eq:uhatbc}
\end{align}
%
Once the solutions to both problems are known, their superposition solves the desired mixed boundary value problem,
%
\begin{align}
     & \left.
    \begin{array}{l}
        \nabla \cdot \tns{\sigma} = \nabla \cdot \left( \tns{\hat{\sigma}} + \tns{\tilde{\sigma}} \right) = \vec{0}                                                             \\
        \tns{\sigma} = \tns{\hat{\sigma}} + \tns{\tilde{\sigma}} = \tns{C} : \left( \tns{\hat{\varepsilon}} + \tns{\tilde{\varepsilon}} \right) = \tns{C} : {\tns{\varepsilon}} \\
        \tns{\varepsilon} = \tns{\hat{\varepsilon}} + \tns{\tilde{\varepsilon}} = \frac{1}{2} \left( \nabla ( \vec{\hat{u}} + \vec{\tilde{u}} ) + \left[ \nabla ( \vec{\hat{u}} + \vec{\tilde{u}} ) \right]^T \right) = \frac{1}{2} \left( \nabla \vec{u} + ( \nabla \vec{u} )^{T} \right)
    \end{array}
    \right\}\quad \textrm{in V}                                    \\
    %
     & \tns{\sigma} \cdot \vec{n} = \vec{T} \quad \textrm{on } S_T \\
    %
     & \left.
    \begin{array}{l}
        \vec{u} = \vec{U}, \quad t>0 \\
        \vec{u} = \vec{0}, \quad t=0
    \end{array}
    \right\} \quad \textrm{on } S_U
    \label{eq:ubc}
\end{align}

For all its simplicity and elegance, the method is not without issue. As the distance between dislocation and surface decreases, $\tns{\tilde{\sigma}}$ diverges \cite{boundary_problems_in_dd}. This can be partially solved by using a non-singular formulation for $\tns{\tilde{\sigma}}$, such as that proposed by \citet{Cai2006}. Regardless, the steep gradients in the dislocation field are difficult to accurately capture as the dislocation approaches $S$. Another problem with this method is the large computational cost when simulating a heterogeneous solid, as this requires calculating polarisation stresses due to the difference in the elastic constants between phases \cite{superposition_scheme0,boundary_problems_in_dd,ddd_precipitate}.

A modified superposition scheme \cite{ODay2004} can overcome this by dividing the problem into separate DDD-FEM problems coupled through a elastic FE problem. This requires accurate evaluation of $\vec{\tilde{T}}$ on the domain boundaries---which can only be captured with a fine FE mesh---increasing the computational cost. Therefore, simulating polycrystalline or composite materials using superposition necesitates methods to accurately evaluate both $\vec{\tilde{u}}$ and $\vec{\tilde{T}}$. The displacements can be evaluated analytically as shown by \cite{bromage2018calculating} and this paper investigates the evaluation of the dislocation tractions analytically.

As previously mentioned, the relative simplicity of the superposition method has made it a popular choice for coupling DDD and FEM as all it needs is the evaluation of FE nodal forces and displacements on the boundary. Furthermore, the analytic expression for the stress field produced by a finite, straight dislocation line segment has allowed \citet{Queyreau} to use the non-singular formulation \cite{Cai2006} to obtain closed-form solutions for the tractions generated by the segment on the surface of a finite element.

The force exerted by a dislocation ensemble on a node, $a$, belonging to element, $e$, is given by,
%
\begin{align}
    \vec{F}_{a} = \int_{S_{e}} \left[\tns{\tilde{\sigma}}(\vec{x}) \cdot \vec{n}\right] N_{a}(\vec{x})\; \mathrm{d}S_{e}\,.
    \label{eq:ddd_fem_force}
\end{align}
%
Where $\mathrm{d}S_{e}$ is the surface of element $e$ with surface area $S_{e}$, and $N_{a}$ is the finite element shape function for node $a$. The solutions are for linear rectangular surface elements and as such the shape functions are,
\begin{align}
    \label{eq:shape_function}
    N_{1} & = \dfrac{1}{4}(1-s_1)(1-s_2)             \\
    N_{2} & = \dfrac{1}{4}(1+s_1)(1-s_2)\nonumber    \\
    N_{3} & = \dfrac{1}{4}(1+s_1)(1+s_2)\nonumber    \\
    N_{4} & = \dfrac{1}{4}(1-s_1)(1+s_2)\nonumber\,.
\end{align}
Where $s_1$ and $s_2$ are the orthogonal coordinates local to the element.

Gauss quadrature is usually used to numerically evaluate the surface integral in \cref{eq:ddd_fem_force}. In 1D this is,
%
\begin{align}
    \label{eq:gauss_leg}
    \int\limits_{-1}^{1} f(s)\;\mathrm{d}s \approx \sum\limits_{i=1}^{n} w_{i} f(s_{i}) \quad \textrm{where}\quad
    %
    w_{i} = \dfrac{2}{\left(1-s_{i}^{2}\right) \left[P_{n}^{'}\left(s_{i}\right)\right]^{2}}\,,
\end{align}
%
is the weighting of the Gauss point, $s_{i}$, which is the $i\textsuperscript{th}$ root of the $n\textsuperscript{th}$ normalised Legendre polynomial, $P_{n}(1) = 1$. $P_{n}^{'}$ is the first order derivative of $P_{n}$. This method is very accurate for functions that can be accurately approximated by polynomials. In fact, for a polynomial of degree $n$, one needs $n-1$ points to obtain an exact numerical solution. However, this quadrature is well-known for being unsuitable for integrating functions with poles or near-poles \cite{gauss_leg, gauss_leg_sing}.

We avoid the strict pole in $\tns{\tilde{\sigma}}$ by using the non-singular formulation described in \cite{Cai2006}, where the true singularity is avoided by adding a small cut-off radius to account for the dislocation core. However, if an integration point falls close to, or within the dislocation core, Gauss quadrature can still produce large errors.

For rectangular surface elements, we must transform from the parent element in $(s_1,\,s_2)$ in \cref{eq:gauss_leg} to the real element coordinate system $(x\,,y\,,z)$. Evaluating \cref{eq:ddd_fem_force} in the parent element and mapping to the real element gives the force on node $a$,
\begin{align}
    \label{eq:gauss_leg_ddd_fem_force}
    \vec{F}_{a} \approx \sum\limits_{i=1}^{Q} w_{i} \sum\limits_{j=1}^{Q} w_{j} [\tns{\tilde{\sigma}}(r_{i}, s_{j})\cdot \vec{n}] N_{a}(r_{i}, s_{j}) \det(\mtx{J})\,.
\end{align}
Where the sum is over the $Q$ quadrature points and $J_{ij} = dx_{i}/ds_{j}$, is the element Jacobian defining the transformation from $(s_1,\,s_2) \mapsto (x,\,y)$. Since the real element is rectangular with surface area $S_e$, then $\det(\mtx{J}) = S_e/4$. \Cref{f:2d_gaussian_quad} contains examples of how the Gauss points are distributed on the surface.
\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{2d_gaussian_quad.pdf}
    \caption{Examples of 2D Gauss-Legendre quadrature of the parent element with $Q = 1,\, 2,\, 3,\, 4$. The point size represents the weight $w$ of the integration point. The parent elements are centred at the origin and $s_1,\, s_2 \in [-1,\,1]$. We use an anticlockwise node numbering scheme.}
    \label{f:2d_gaussian_quad}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.8\linewidth]{force_calc_linear_rectangle.pdf}
    \caption{Diagram of the parametric line integrals solved by \citet{Queyreau} to find the forces on linear rectangular surface elements.}
    \label{f:force_lin_rect}
\end{figure}
Explicitly, \cref{eq:ddd_fem_force} is actually a triple vector integral as shown in \cref{f:force_lin_rect}. This is because given the isotropic Burgers vector distribution proposed in \cite{Cai2006}, the dyadic form of the stress tensor produced by a straight, finite dislocation segment bounded by nodes at $\mathbf{x_1}$ and $\mathbf{x_2}$ \cite{Queyreau} is,
\begin{align}
    \label{eq:stress}
    \tns{\tilde{\sigma}}(\vec{x}) = &
    - \dfrac{\mu}{8\pi} \int\limits_{\mathbf{x_1}}^{\mathbf{x_2}} \left( \dfrac{2}{R_{a}^{3}} + \dfrac{3a^2}{R_{a}^{5}} \right) \left[ \left(\vec{R} \times \vec{b}\right) \otimes \mathrm{d}\vec{x'} + \mathrm{d}\vec{x'} \otimes \left(\vec{R} \times \vec{b}\right) \right]          \\
    %
                                    & + \dfrac{\mu}{4\pi(1-\nu)} \int\limits_{\mathbf{x_1}}^{\mathbf{x_2}} \left( \dfrac{1}{R_{a}^{3}} + \dfrac{3a^2}{R_{a}^{5}} \right) \left[ \left(\vec{R} \times \vec{b}\right) \cdot \mathrm{d}\vec{x'} \right]\vec{I_2}\nonumber                  \\
    %
                                    & -\dfrac{\mu}{4\pi(1-\nu)} \int\limits_{\mathbf{x_1}}^{\mathbf{x_2}}  \dfrac{1}{R_{a}^{3}} \left[ \left(\vec{b} \times \mathrm{d}\vec{x'}\right) \otimes \vec{R} + \vec{R} \otimes \left(\vec{b} \times \mathrm{d}\vec{x'}\right) \right]\nonumber \\
    %
                                    & + \dfrac{\mu}{4\pi(1-\nu)} \int\limits_{\mathbf{x_1}}^{\mathbf{x_2}} \dfrac{3}{R_{a}^{5}} \left[ \left(\vec{R} \times \vec{b}\right) \cdot \mathrm{d}\vec{x'} \right]\vec{R}\otimes\vec{R}\nonumber,
\end{align}
where,
\begin{align}
    \vec{R}            & = \vec{x} - \vec{x'} = y \vec{l} + r \vec{p} + s \vec{q} \\
    %
    R_a                & = \sqrt{\vec{R} \cdot \vec{R} + a^2}                     \\
    %
    \mathrm{d}\vec{x'} & = -\mathrm{d} y \vec{l}\,.
\end{align}
The vectors $\vec{p}$ and $\vec{q}$ are aligned with the edges of the rectangular finite element, $\vec{n} = \vec{p} \times \vec{q}$ is the element surface normal (pointing away from the dislocation), and $\vec{l}$ is parallel to the dislocation line segment as shown in \cref{f:force_lin_rect}. Then (provided $\vec{l}$ is not parallel to $\vec{p}$ or $\vec{q}$) $\vec{R}$ can be expressed in terms of $(\vec{l},~\vec{p},~\vec{q})$ with coefficients,
\begin{align}
    y = \dfrac{\vec{R}\cdot \vec{n}}{\vec{l}\cdot \vec{n}} \label{eq:problem},\quad
    %
    r = \dfrac{\vec{R}\cdot (\vec{q} \times \vec{l})}{\vec{p}\cdot (\vec{q} \times \vec{l})}, \quad
    %
    s = \dfrac{\vec{R}\cdot (\vec{p} \times \vec{l})}{\vec{q}\cdot (\vec{p} \times \vec{l})}\,.
\end{align}

Substituting \cref{eq:stress} and \cref{eq:shape_function} into \cref{eq:ddd_fem_force} yields four long and messy equations (one for each FE node) that were elegantly solved by \citet{Queyreau} by utilising the fact that the triple integrals all had the form,
\begin{align}
    H_{ijkl}        & = \int\limits_{r_{1}}^{r_{2}}\int\limits_{s_{1}}^{s_{2}}\int\limits_{y_{1}}^{y_{2}} \dfrac{r^i s^j y^k}{R_{a}^{m}}\label{eq:triple_int} \\
    \textrm{when }m & = 5 \textrm{ then } i,\, j \in [0,\,3],~k \in [0,\,2]\nonumber                                                                          \\
    \textrm{when }m & = 3 \textrm{ then } i,\, j \in [0,\,2],~k \in [0,\,1]\nonumber                                                                          \\
    \textrm{when }m & = 1 \textrm{ then } i = j = k = 0\,.
\end{align}
Using partial differentiation and integration by parts, they found a series of recurrence relations that lead to double and single integrals of similar form to \cref{eq:triple_int}. All of which are used to construct a full, exact solution. The recurrence relations stop working when $i = j = k = 0 \textrm{ and } m = 1,\, 3$. At which point, direct integration of the remaining single and double integrals (the last triple integrals all cancel out in the global calculation) yields six seed functions that are used as the starting point for the recurrence relations. Three of them are logarithms and three either arctangents or---if a discriminant is negative---hyperbolic arctangents. The details of the procedure can be found in \cite{Queyreau}.

Although exact, the use of arctangents, hyperbolic arctangents and logarithmic functions, compounded by the large number of recurrence relations is prime territory for error propagation and numerical problems (see \cref{s:method}). The problem is particularly egregious when using general purpose compilers instead of high-performance or scientific computing compilers where mathematical functions are implemented more precisely. Such issues must be taken into account when using analytic tractions, which can be done by using numerical tolerances as described in \cref{s:method}.

In simulations, tractions are manifested as image stresses calculated by the FE solver at FE nodes. In order to validate and compare the practical differences between analytic and numeric methods, we compare the resulting image stresses from both methods to the analytic expressions for infinite dislocations in inhomogenous media for edge dislocations \cite{head1953edge}, as well as those for screw dislocations \cite[p.~59,~64]{hirth1983theory}. We keep the same nomenclature and coordinate system as both infinite-domain solutions. Where the traction surface is the line $x=0$, the dislocation line direction is the positive $z$-direction (pointing out of the page), the dislocation coordinates are represented by $(a,~c)$, and points in the $xy$-plane described by their $(x,\,y)$ coordinates.

The original paper by \citet{head1953edge} has a few typos that have been replicated in other sources, and the equations in \cite[p.~59,~64]{hirth1983theory} assume the dislocation is found at the origin. We therefore include the complete and correct expressions in \cref{eq:imageStressAnalyticEdge1,eq:imageStressAnalyticEdge2,eq:imageStressAnalyticScrew}. \citet{head1953edge} gives two basic cases. \Cref{eq:imageStressAnalyticEdge1} corresponds to the case where $\vec{b}$ is perpendicular to the surface and positive $b$ means it points in the positive $x$-direction,
\begin{subequations}\label{eq:imageStressAnalyticEdge1}

@@

        \sigma_{xx} = D (y - c) & \left\{-\dfrac{3 (x - a)^2 + (y - c)^2}{[(x - a)^2 + (y - c)^2]^2} + \dfrac{3 (x + a)^2 + (y - c)^2}{[(x + a)^2 + (y - c)^2]^2}\right.                                             \\\nonumber
                                & \left. + 4 a x \dfrac{3 (x + a)^2 - (y - c)^2}{[(x + a)^2 + (y - c)^2]^3}\right\}\,,                                                                                               \\
        \sigma_{yy} = D (y - c) & \left\{\dfrac{(x - a)^2 - (y - c)^2}{[(x - a)^2 + (y - c)^2]^2} - \dfrac{(x + a)^2 - (y - c)^2}{[(x + a)^2 + (y - c)^2]^2}                                                 \right. \\\nonumber
                                & \left. + 4 a (2 a - x) \dfrac{(x + a)^2 + (3 x + 2 a) (y - c)^2}{[(x + a)^2 + (y - c)^2]^3}\right\}\,,                                                                             \\
        \sigma_{xy} = D         & \left\{(x - a) \dfrac{(x - a)^2 - (y - c)^2}{[(x - a)^2 + (y - c)^2]^2} - (x + a) \dfrac{(x + a)^2 - (y - c)^2}{[(x + a)^2 + (y - c)^2]^2}\right.                                  \\\nonumber
                                & \left. + 2 a \dfrac{6 x (x + a) (y - c)^2 - (x - a) (x + a)^3 - (y - c)^4}{[(x + a)^2 + (y - c)^2]^3}\right\}\,.

@@

\Cref{eq:imageStressAnalyticEdge2} corresponds to the case where the $\vec{b}$ lies parallel to the surface and positive $b$ means it points in the positive $y$-direction,
\begin{subequations}\label{eq:imageStressAnalyticEdge2}

@@

        \sigma_{xx} = D         & \left\{ (x - a) \dfrac{(x - a)^2 - (y - c)^2}{[(x - a)^2 + (y - c)^2]^2} -(x + a) \dfrac{(x + a)^2 - (y - c)^2}{[(x + a)^2 + (y - c)^2]^2} \right.     \\\nonumber
                                & \left. + 2 a \dfrac{(3 x + a) (x + a)^3 - 6 x (x + a) (y - c)^2 - (y - c)^4}{[(x + a)^2 + (y - c)^2]^3}\right\}\,,                                     \\
        \sigma_{yy} = D         & \left\{ (x - a) \dfrac{(x - a)^2 + 3 (y - c)^2}{[(x - a)^2 + (y - c)^2]^2} - (x + a) \dfrac{(x + a)^2 + 3 (y - c)^2}{[(x + a)^2 + (y - c)^2]^2}\right. \\\nonumber
                                & \left. - 2 a \dfrac{(x - a) (x + a)^3 - 6 x (x + a) (y - c)^2 + (y - c)^4}{[(x + a)^2 + (y - c)^2]^3}\right\}\,,                                       \\
        \sigma_{xy} = D (y - c) & \left\{\dfrac{(x - a)^2 - (y - c)^2}{[(x - a)^2 + (y - c)^2]^2} - \dfrac{(x + a)^2 - (y - c)^2}{[(x + a)^2 + (y - c)^2]^2} \right.                     \\\nonumber
                                & \left. + 4 a x \dfrac{3 (x + a)^2 - (y - c)^2}{[(x + a)^2 + (y - c)^2]^3}\right\}\,.

@@

\Cref{eq:imageStressAnalyticScrew} corresponds to screw dislocations, which are markedly simpler as only the shear components are non-zero. Here $\vec{b} = \vec{l}$ so positive $b$ means it points in the positive $z$-direction,
\begin{subequations}\label{eq:imageStressAnalyticScrew}

@@

        \sigma_{xz} & = -D \left(\dfrac{y - c}{(x - a)^2 + (y - c)^2} - \dfrac{y - c}{(x + a)^2 + (y - c)^2}\right)   \\
        \sigma_{yz} & = D \left(\dfrac{x - a}{(x - a)^2 + (y - c)^2} - \dfrac{x + a}{(x + a)^2 + (y - c)^2}\right)\,.

@@

In every case, the constant $D$ is defined by \cref{eq:Dconstant},
\begin{align}\label{eq:Dconstant}
    D = \dfrac{\mu}{2\pi} \cdot \dfrac{1+\nu}{1-\nu^2} \cdot b\,.

@@

Note the first terms of \cref{eq:imageStressAnalyticEdge1,eq:imageStressAnalyticEdge2,eq:imageStressAnalyticScrew} all correspond to the stress field generated by the dislocation itself. The following terms are the corrective terms required to make the boundary conditions on the surface equal to zero. Therefore, we can split these equations and only look at the real or corrective terms independently, which we do in order to only visualise the effect of the different traction calculations. Furthermore, \cref{eq:imageStressAnalyticEdge1,eq:imageStressAnalyticEdge2,eq:imageStressAnalyticScrew} are all singular at the dislocation coordinates. Our simulation code uses the non singular expressions found by \citet{Cai2006}, which smooth out drastic increases in stresses and avoid numerical blow up as we near the dislocation core.

\section{Methodology}\label{s:method}
